Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-10 paper/paper.tex
conclusions ...

%% One issue we  highlighted was that while fewer  nodes were marked (and
%% more garbage collected), sometimes  cons cells could be visited and
%% traversed  multiple times  with different  sets of  liveness  paths to
%% explore.  Further work includes  improvements to the classical copying
%% collector to reduce the cost of this.
{\color{blue}In our current  work we have extended  the liveness based
  garbage collection to lazy languages  and show that it has benefits.
  Our  biggest   contribution  is  our  ability   to  handle  closures
  (suspended evaluations) during garbage collection.  We show that for
  a  practical  garbage collector  it  is  not possible  to  implement
  liveness based collector for  closures efficiently.  By implementing
  a mixed-mode garbage collection scheme, we were able to collect more
  cells than reachability based  collectors without paying the penalty
  of  a full  liveness based  collector.  We  believe this  is a  very
  practical idea and hence easily implementable.

We  also overcome  the drawback  of  multiple traversals  that a  full
liveness based  collector has  by using  an extra  bit and  setting it
whenever  the  cell   is  copied  using  reachability(as   part  of  a
closure). Our experimental results  have shown that such optimizations
provide performance benefits with very little loss of precision. Using
the  liveness  annotations  for  the  print  primitive  also  improves
performance and subsumes compile  time optimizations such as tail-call
optimization.

%%\subsection{Future work}
Although we  provide an  implementation for  a liveness  based garbage
collection scheme for lazy language, we  do not provide a formal proof
of its correctness.  A formal  proof of correctness would describe how
closures  are  handled  during  garbage  collection  and  liveness  is
propogated inside closures.

 Another interesting exercise is to narrow
the gap between the actual liveness  and the perceived liveness of our
garbage collector. Our  experiments show that a  significant number of
dead cells get trapped inside  closures. Using strictness analysis and
eagerly evaluating closures might release  more of these dead cells to
be  garbage  collected.   Since   our  garbage  collector  can  handle
closures(suspended evaluations)  another challenge would be  to extend
it to support higher order programs.

Orthogonally,  we can  improve the  efficiency of  the liveness  based
garbage  collector so  that  even  when the  decrease  in the  garbage
collections compared  to reachability is  not much, the  total garbage
collection is  not bad. Making  the DFA  smaller by using  better data
structures like Binary Decision Diagrams(BDD), truncating the liveness
automata  after   certain  depth(k-liveness)  are  some   ideas  worth
investigating.

>>>>

%% One issue we  highlighted was that while fewer  nodes were marked (and
%% more garbage collected), sometimes  cons cells could be visited and
%% traversed  multiple times  with different  sets of  liveness  paths to
%% explore.  Further work includes  improvements to the classical copying
%% collector to reduce the cost of this.
{\color{blue}We extended the liveness based garbage collection to lazy
  languages and shown its benefit  for garbage collection in practical
  programs. We  defined context  senstive liveness analysis  that uses
  context independent summaries of functions obtained using a symbolic
  demand. We show  that obtaining precise solution  to these equations
  is undecidable in  our formulation, and hence  we safely approximate
  the result using DFAs. These DFAs are consulted by garbage collector
  to improve to collection.

  Liveness   of   closure   presented  further   challenges   to   our
  implementation.   We compared  different approximate  strategies for
  handling   liveness   of   closures,   and  found   that   a   mixed
  strategy---reachability  based  collection   of  closure  arguments,
  liveness  based  collection  of   everything  else---works  best  in
  practice as it  avoid runtime and space  overheads.  With mixed-mode
  garbage collection scheme,  we were able to collect  more cells than
  reachability based collectors at a reasonable overhead.

  Although we  provide an  implementation for  a liveness  based garbage
  collection scheme for lazy language, we  do not provide a formal proof
  of its correctness.  A formal  proof of correctness would describe how
  closures  are  handled  during  garbage  collection  and  liveness  is
  propogated inside closures.

  Another interesting exercise is to narrow the gap between the actual
  liveness and  the perceived liveness  of our garbage  collector. Our
  experiments (not  reported here) show  that a significant  number of
  dead cells  get trapped inside closures.   Using strictness analysis
  to eagerly evaluate closures might  release more of these dead cells
  to be  garbage collected.   Since our  garbage collector  can handle
  closures  (suspended evaluations),  another  major  challenge is  to
  extend it to support higher order programs.

  Orthogonally,  we plan  to improve  the efficiency  of the  liveness
  based garbage collector using heuristics  such as limiting the depth
  of  DFAs,   merging  nearly-equivalent   states  and   using  better
  representation  and algorithms  for automata  manipulation. We  also
  need  to   investigate  the  interaction  of   liveness  with  other
  collection schemes, such as incremental and generational collection.
  In short,  we need to  investigate the  ways to make  liveness based
  garbage collection attractive for practical collectors.  }
